### 2023-01-10 tex/framework.tex
[ writing ] renew generic syntax chapter

\chapter{A framework for usage-restricted calculi}\label{sec:framework}

@@

In this chapter, I take the framework for typed syntaxes with binding developed
by \citet{AACMM21} and apply the principles we discovered in
\cref{sec:semirings} to yield a framework allowing for semiring-based usage
restrictions on variables.
Syntactically, I claim that this framework ranges over all finitiary
variable-based simply typed semiring-annotated calculi, with justification by
comparison to the framework of \citet{AACMM21} and some novel examples in
\cref{sec:other-syntaxes}.
I also derive analogues to some of the semantic results of \citet{AACMM21},
strengthening them to take advantage of usage restrictions.

The work in this chapter is fully mechanised in Agda, which allows me to be
precise about the various levels of domain-specific languages which appear.
I assume that the reader is familiar with the bunched connectives introduced in
\cref{fig:bunched} and the usage-aware environments of \cref{def:lr-env}.

\section{Syntax}

@@

\section{Semantics}

>>>>

\chapter{Generic usage-annotated syntax}\label{sec:framework}

@@

In this chapter, I take the framework for typed syntaxes with binding developed
by \citet{AACMM21} and apply the principles we discovered in
\cref{sec:semirings} to yield a framework for defining calculi with
semiring-based usage restrictions on variables.
Syntactically, I claim that this framework ranges over all finitiary
variable-based simply typed semiring-annotated calculi, with justification by
comparison to the framework of \citet{AACMM21} and some novel examples in
\cref{sec:other-syntaxes}.
%I also derive analogues to some of the semantic results of \citet{AACMM21},
%strengthening them to take advantage of usage restrictions.

The work in this chapter is fully mechanised in Agda, which allows me to be
precise about the various levels of domain-specific languages which appear.

@@

\chapter{Generic usage-aware semantics}\label{sec:semantics}

### 2023-01-10 tex/semantics.tex
[ edit ] some corrections and notes after meeting
[ writing ] renew generic syntax chapter

Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
\AgdaFunction{semantics} we define in this section assigns a
$\C$-value in context $\Gamma$ to every term in context $\Delta$,
where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
semantic interpretation of terms ($\V$ being the semantic
interpretation of variables). Before we can define
\AgdaFunction{semantics}, we need to treat recursion through rules'
premises (\cref{sec:functorial}) and extension of environments when
going under variable binders (\cref{sec:kripke}).

@@

\subsection{A layer of syntax is functorial}\label{sec:functorial}

A basic property of the universe of syntaxes we described in \cref{sec:syntax}
is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{F}.

\ExecuteMetaData[\Maptex]{map-s-type}

This generality is needed because usage contexts change between
a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

\ExecuteMetaData[\Maptex]{map-p}

\subsection{The Kripke function space}\label{sec:kripke}

@@

\subsection{Semantic traversal}\label{sec:traversal}

>>>>

Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
\AgdaFunction{semantics} we define in this section assigns a
$\C$-value in context $\Gamma$ to every term in context $\Delta$,
where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
semantic interpretation of terms ($\V$ being the semantic
interpretation of variables). Before we can define
\AgdaFunction{semantics}, we need to treat recursion through rules'
premises (\cref{sec:functorial}) and extension of environments when
going under variable binders (\cref{sec:kripke}).
\todo{Introduction for the chapter}

@@

\section{The Kripke function space}\label{sec:kripke}

@@

\section{Semantic traversal}\label{sec:traversal}

### 2023-01-10 tex/syntax.tex
[ writing ] renew generic syntax chapter

The aim of this section is to produce a domain-specific language of
\emph{syntax descriptions}.
One may also use the terminology ``universe'' in place of ``domain-specific
language'', alluding to the \emph{universe pattern} common in dependently typed
programming~\citep{BDJ03}.
A syntax description is essentially a concise, high-level representation of a
type system's syntactic rules.
The information contained in a syntax description is comparable to what is
written (informally) in \cref{fig:lr-bunched}.
The key features allowing these descriptions to capture semiring-annotated
calculi are the distinction between sharing ($\dottimes$) and separating ($*$)
conjunction of premises, modal scaling by a semiring element ($\cdot$), and the
inclusion of semiring annotations on newly bound variables.

@@

\subsection{Descriptions of Systems}

@@

A \AgdaRecord{Rule} is a pair of some \AgdaDatatype{Premises} and a
conclusion. We use an infix arrow as a suggestive notation for rules.

@@

\noindent
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\PaperExamplestex]{Ty}
  \ExecuteMetaData[\PaperExamplestex]{Side}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\PaperExamplestex]{qlR}
\end{minipage}

@@

\subsection{Terms of a \AgdaRecord{System}}\label{sec:terms}

@@

\subsection{Other syntaxes and syntactic forms}\label{sec:other-syntaxes}

\paragraph{The system $\mu\tilde\mu$.}
We can encode a usage-annotated version of System $L$/the
$\mu\tilde\mu$-calculus~\cite{CH00} --- a syntax for classical logic --- in
such a way that contexts capture the undistinguished parts of the sequent.
As such, the generic substitution lemma we get in \cref{sec:kit-to-sem} is the
form of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
Though the $\mu\tilde\mu$-calculus is originally described as a sequent
calculus~\cite{CH00}, we use the techniques of
\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it as a natural
deduction system, thus giving a notion of \emph{variable} to the system.

Unlike the single judgement form of \name{} and standard simply typed
$\lambda$-calculi, the $\mu\tilde\mu$-calculus has three judgement forms:
terms, coterms, and commands.
Read logically, terms and coterms are seen to, respectively, prove and refute
propositions (types), while commands exhibit contradictions.
This means that the abstract \AgdaBound{Ty} in the generic framework is
instantiated to \AgdaDatatype{Conc} (for \emph{conclusion}) as below, with
\AgdaDatatype{Ty} not being exposed directly to the generic framework.
For now, we just consider multiplicative disjunction $\parr$ (\emph{par}) and
negation/duality, beside an uninterpreted base type.
These are enough to exhibit classical behaviour.

@@

\paragraph{Duplicability}
There is one more bunched combinator we have experimented with adding to the
framework:

\[
  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
  \times \plr{\grRprime \leq \gr0}
  \times \plr{\grRprime \leq \grRprime + \grRprime}
  \times T\,\grRprime
\]

The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
of it, can be both discarded and duplicated indefinitely, and in the
refinement we have a $T$.
We use this combinator to introduce subterms that are used an unknown number of
times, for example the continuations of the eliminator of an inductive type,
or other fixed points.
We can also use it in linear/non-linear style systems~\cite{Benton94} to make
sure linear variables are not available in the intuitionistic fragment.

Adding the $\Box$ combinator is the only thing we have found that requires our
linear maps be functional rather than merely relational.

>>>>

The aim of this chapter is to produce a domain-specific language of
\emph{syntax descriptions}.
One may also use the terminology ``universe'' in place of ``domain-specific
language'', alluding to the \emph{universe pattern} common in dependently typed
programming~\citep{BDJ03}.
A syntax description is essentially a concise, high-level representation of a
type system's syntactic rules.
The information contained in a syntax description is comparable to what is
written (informally) in \cref{fig:lr-bunched}.
The key features allowing these descriptions to capture semiring-annotated
calculi are the distinction between sharing ($\dottimes$) and separating ($*$)
conjunction of premises, modal scaling by a semiring element ($\cdot$), and the
inclusion of semiring annotations on newly bound variables.

@@

\section{Descriptions of Systems}

@@

A \AgdaRecord{Rule} is a pair of some \AgdaDatatype{Premises} and a
conclusion.
I suggestively use a quoted version of the ``universal entailment'' arrow
$\rightarrowtriangle$, the unquoted version of which interprets the horizontal
line in a traditionally presented typing rule.

@@

\noindent
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\PaperExamplestex]{Ty}
  \ExecuteMetaData[\Handtex]{Hand}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \ExecuteMetaData[\PaperExamplestex]{qlR}
\end{minipage}

@@

\section{Terms of a \AgdaRecord{System}}\label{sec:terms}

@@

\section{More example syntaxes}\label{sec:other-syntaxes}

With the range of representable syntaxes now formalised, we can explore encoding
techniques for syntaxes more exotic than ST$\lambda$C and $\name$.
As well as the variations presented in \cref{sec:variant}, we can represent a
usage-annotated $\mu\tilde\mu$-calculus and a Linear/non Linear system.

\subsection{The system $\mu\tilde\mu$}
I encode a usage-annotated version of System $L$/the
$\mu\tilde\mu$-calculus~\citep{CH00} --- a syntax for classical logic --- in
such a way that contexts capture the undistinguished parts of the sequent.
As such, the generic substitution lemma we get in \cref{sec:kit-to-sem} is the
form of substitution required in standard $\mu\tilde\mu$-calculus metatheory.
Though the $\mu\tilde\mu$-calculus is originally described as a sequent
calculus~\citep{CH00}, I use the techniques of
\citet[p.~12]{herbelin-hab} and \citet{LC06} to present it using hypothetical
judgements, thus giving a notion of \emph{variable} to the system.

Unlike the single judgement form of \name{} and standard simply typed
$\lambda$-calculi, the $\mu\tilde\mu$-calculus has three judgement forms:
terms, coterms, and commands.
Read logically, terms and coterms are seen to, respectively, prove and refute
propositions (types), while commands exhibit contradictions.
This means that the abstract \AgdaBound{Ty} in the generic framework is
instantiated to \AgdaDatatype{Conc} (for \emph{conclusion}) as below, with
\AgdaDatatype{Ty} not being exposed directly to the generic framework.
For now, I just consider multiplicative disjunction $\parr$ (\emph{par}) and
negation/duality, beside an uninterpreted base type.
These are enough to exhibit classical behaviour.

@@

\subsection{Duplicability and L/nL}\label{sec:dup-lnl}
In \cref{sec:rec}, I introduced a bunched connective $\Box^{0{+}}$, used to
ensure that such a premise is derived in a context that allows it to be
duplicated and discarded.
Syntax for this connective can be added to \AgdaDatatype{Premises} and worked
through the various parts of the framework, though I do not show these parts in
the text of this thesis for the sake of simplicity and brevity.
Additionally, variations such as $\Box^0$ (requiring only that the context is
${} \leq \gr0$) and $\Box^{0{+}{*}}$ (requiring, in addition to
$\Box^{0{+}}$, closure under multiplication by any scalar) can be added to the
framework independently of each other and of $\Box^{0{+}}$.

As well as the use case of inductive types, as shown in \cref{sec:rec}, I use
the $\Box^{0{+}{*}}$-modality to encode the restriction to intuitionistic
variables in the intuitionistic judgement forms of the L/nL
calculus~\citep{Benton94}.
I will explain this in detail in due course.

\todo{Maybe partially merge with the later L/nL section.}
Linear/non-Linear Logic has two sorts of types: linear types $A, B, C$ and
intuitionistic types $X, Y, Z$.
I encode these two sorts in the Agda types
\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{lin} and
\AgdaDatatype{Ty}\AgdaSpace{}\AgdaInductiveConstructor{int}, respectively.
Using an indexed type lets me take the total space of \AgdaDatatype{Ty}, named
\AgdaFunction{$\Sigma$Ty}, as the evident dependent pair type.
\AgdaFunction{$\Sigma$Ty} becomes the type of types seen by the framework.
The types themselves are, on the linear side,
a tensor unit \AgdaInductiveConstructor{tI},
a tensor product \AgdaInductiveConstructor{\_t$\otimes$\_}, and
a linear function space \AgdaInductiveConstructor{\_t$\multimap$\_};
on the intuitionistic side,
a unit \AgdaInductiveConstructor{t1},
a product \AgdaInductiveConstructor{\_t$\times$\_}, and
an intuitionistic function space \AgdaInductiveConstructor{\_t$\to$\_};
and adjoint type formers \AgdaInductiveConstructor{tF} and
\AgdaInductiveConstructor{tG}.
I also throw in a linear base type \AgdaInductiveConstructor{$\iota$}.

\ExecuteMetaData[\LnLtex]{Frag}
\ExecuteMetaData[\LnLtex]{Ty}
%\ExecuteMetaData[\LnLtex]{SigTy}

As in the other examples, I define a data type of rule labels
\AgdaDatatype{`LnL} before defining the type system \AgdaFunction{LnL}.
The system \AgdaFunction{LnL} has types \AgdaFunction{$\Sigma$Ty} and usage
annotations $\{\gr0, \gr1, \gr\omega\}$, as we have used before for linear
systems.

We want to be able to embed the L/nL judgement forms
$\Theta; \Gamma \vdash_{\mathcal L} A$ and $\Theta \vdash_{\mathcal C} X$, where
$\Theta$ is made of intuitionistic variables and $\Gamma$ is made of linear
variables.
Following allowed usage patterns, intuitionistic variables will all have usage
annotation $\gr\omega$, while linear variables will each be annotated
either $\gr1$ or $\gr0$.
These assignments of usage annotations are enforced wherever the rules bind new
variables.
By usage subsumption, intuitionistic variables may also end up with annotation
$\gr1$ or $\gr0$, but we see this as a (meaningless) refinement of the
$\gr\omega$ annotation, not causing a soundness problem.
The key invariant is that linear variables never get annotation $\gr\omega$.

The framework already tells us what variables and contexts we can
have --- namely that we can make any assumption in any context.
This means that we can propose ill formed sequents like
$\gr1A \vdash X$.
To make sure that this is not a problem, we want to maintain the invariant that
any \emph{derivation} of a well formed sequent contains only well formed
sequents.
I enforce this locally using the $\Box^{0{+}{*}}$ bunched modality:
Every rule targeting an intuitionistic judgement has its premises wrapped in a
$\Box^{0{+}{*}}$ to restrict the usage annotations to $\gr0$ and $\gr\omega$.
Additionally, rules \TirName{$F$i} and \TirName{$G$e}, which target linear
judgements but have intuitionistic premises, similarly get a $\Box^{0{+}{*}}$.
Note that having the annotation $\gr0$ means that well formed intuitionistic
sequents can contain linear variables annotated $\gr0$, but these variables are
not usable.

The rules are presented in Agda code below, and also using bunched inference
rule notation in \cref{fig:LnL-bunched}

\ExecuteMetaData[\LnLtex]{quoteLnL}
\ExecuteMetaData[\LnLtex]{LnL}

\begin{figure}
  \begin{mathpar}
    \ebrule[comb]{%
      \hypo{I^{\sep}}
      \infer1[$I$i]{\vdashL I}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL I}
      \hypo{\sep}
      \hypo{\vdashL C}
      \infer3[$I$e]{\vdashL C}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL A}
      \hypo{\sep}
      \hypo{\vdashL B}
      \infer3[$\otimes$i]{\vdashL A \otimes B}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL A \otimes B}
      \hypo{\sep}
      \hypo{\gr1A, \gr1B \vdashL C}
      \infer3[$\otimes$e]{\vdashL C}
    }
    \and
    \ebrule[comb]{%
      \hypo{\gr1A \vdashL B}
      \infer1[$\multimap$i]{\vdashL A \multimap B}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL A \multimap B}
      \hypo{\sep}
      \hypo{\vdashL A}
      \infer3[$\multimap$e]{\vdashL B}
    }
    \and
    \ebrule[comb]{%
      \hypo{\Boxzpt\plr{\vdashC X}}
      \infer1[$F$i]{\vdashL FX}
    }
    \and
    \ebrule[comb]{%
      \hypo{\vdashL FX}
      \hypo{\sep}
      \hypo{\gr\omega X \vdashL C}
      \infer3[$F$e]{\vdashL C}
    }
    \and
    \ebrule[comb]{%
      \hypo{\Boxzpt\dot1}
      \infer1[$1$i]{\vdashC 1}
    }
    \and
    \text{(no \TirName{$1$e})}
    \and
    \ebrule[comb]{%
      \hypo{\Boxzpt(\vdashC X}
      \hypo{\dottimes}
      \hypo{\vdashC Y)}
      \infer3[$\times$i]{\vdashC X \times Y}
    }
  \end{mathpar}
  \caption{Linear/non-Linear Logic in bunched notation}\label{fig:LnL-bunched}
\end{figure}

Later,\todo{When?} I will give translations between this encoding of L/nL and
$\name$, providing some assurance that the correct logic has been encoded.

%There is one more bunched combinator we have experimented with adding to the
%framework:
%
%\[
%  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
%  \times \plr{\grRprime \leq \gr0}
%  \times \plr{\grRprime \leq \grRprime + \grRprime}
%  \times T\,\grRprime
%\]
%
%The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
%of it, can be both discarded and duplicated indefinitely, and in the
%refinement we have a $T$.
%We use this combinator to introduce subterms that are used an unknown number of
%times, for example the continuations of the eliminator of an inductive type,
%or other fixed points.
%We can also use it in linear/non-linear style systems~\cite{Benton94} to make
%sure linear variables are not available in the intuitionistic fragment.
%
%Adding the $\Box$ combinator is the only thing we have found that requires our
%linear maps be functional rather than merely relational.

\section{A layer of syntax is functorial}\label{sec:functorial}

A basic property of the universe of syntaxes
is that every syntax supports a functorial action on subterms, realised by a
function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{$\Psi$}.
The notation
\AgdaBound{$\Psi$}\AgdaSpace{}\AgdaSymbol{.}\AgdaField{rel}\AgdaSpace{}%
\AgdaBound{P}\AgdaSpace{}\AgdaBound{Q}
means the same as $\grQ \leq \grP\gr\Psi$ from \cref{sec:ren-sub-lr}.
The new notation comes about because I state \AgdaBound{$\Psi$} as a total and
deterministic linear relation, rather than a linear map, for reasons related to
mechanisation beyond the scope of this thesis.

\ExecuteMetaData[\Maptex]{map-s-type}

This generality is needed because usage contexts change between
a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{$\Psi$} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

The linearity of relation \AgdaBound{$\Psi$} is given by fields
\AgdaField{rel-0$_m$}, \AgdaField{rel-+$_m$}, and \AgdaField{rel-*$_m$} (with
the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
\todo{Refer to preliminaries}

\ExecuteMetaData[\Maptex]{map-p}

I have also extended \AgdaFunction{map-p} to handle the various
$\Box$-modalities described in \cref{sec:dup-lnl}.
The Agda code for this extension is not particularly readable, so I do not
include it in this document.
However, this extension is notable as the only part of the framework requiring
that the linear relation \AgdaBound{$\Psi$} be functional (i.e., total and
deterministic).
